### 2023-10-14 chapters/paper_benchmarking_workshop.tex
Some work on discussion
Updates to discussion
Fix many spelling and language errors

        \edge[]{d_t}{z_t};
        \edge[bend right]{d_t}{x_t};
        \edge[]{z_tm1}{d_t};
        \edge[]{d_tm1}{d_t};
        \edge[]{z_t}{x_t};
    \end{tikzpicture}
    }
    \end{multicols}
\caption{
Generative models of an LSTM, VRNN, SRNN, STCN and CW-VAE for a single time step. The STCN and CW-VAE are illustrated with a single latent variable. 
Red arrows indicate purely deterministic paths from the output $\vx_t$ to previous input $\vx_{<t}$ without passing a stochastic node. 
We provide additional graphical illustrations including inference models in \cref{app: additional graphical models}.
}
\label{fig: autoregressive vrnn srnn cwvae 1L graphs}
\end{figure*}

\paragraph{Scope and related work} 
At a high level, this benchmark brings order to LVM model comparisons for speech and also provides useful reference implementations of the models\footnote{\texttt{ \href{https://github.com/JakobHavtorn/benchmarking-lvms}{github.com/JakobHavtorn/benchmarking-lvms}}}. 
An extended version of this paper is available at \href{https://arxiv.org/abs/2202.12707}{arXiv:2202.12707}.
LVMs based on the VAE are of interest due to their ability to learn an approximate posterior distribution over latent variables. This makes them useful beyond generation for e.g. semi"=supervised learning \parencite{kingma_semi"=supervised_2014} and anomaly detection \parencite{havtorn_hierarchical_2021}. 
In this paper, we focus on the VRNN \parencite{chung_recurrent_2015}, SRNN \parencite{fraccaro_sequential_2016} and STCN \parencite{aksan_stcn_2019} since they are similar to the original VAE framework. 
We exclude some related work from the benchmark. Specifically, the FH-VAE \parencite{hsu_unsupervised_2017} adds a global latent variable but we exclude it due to it's segmentation of the training data. We also exclude Z-forcing \parencite{goyal_zforcing_2017} which resembles the SRNN, but deviates from maximum likelihood training by using an auxiliary loss. We exclude the VQ-VAE \parencite{oord_neural_2018} which is a hybrid LVM and autoregressive model, since it uses a quantized latent space. Finally, we exclude the Stochastic WaveNet \parencite{lai_stochastic_2018} which is similar, but inferior, to the STCN \parencite{aksan_stcn_2019}.

@@

\section{Models}
\paragraph{Sequential deep latent variable models}
The models we consider are all sequential deep latent variable models trained with variational inference and the reparameterization trick \parencite{kingma_autoencoding_2014}.
The input is a variable-length sequence $\vx = \vx_{1:T} = \left(\vx_1, \vx_2, \dots, \vx_{T}\right)$ with $\vx_t\in\mathbb{R}^{D_x}$. 
The models first encode $\vx_{1:T}$ to a temporal posterior distribution $q(\vz|\vx)$ and sample a latent representation $\vz_{1:T}$ where $\vz_t\in\mathbb{R}^{D_z}$. This is then used to reconstruct the input while the posterior distribution is regularized to be close to a prior distribution $p(\vz_t|\cdot)$ where the dot indicates that it may depend on latent and observed variables at previous time steps, $\vz_{<t}$ and $\vx_{<t}$ where $\vz_{<t} := \left( \vz_1, \vz_2, \dots, \vz_{t-1} \right)$. 
The models are trained to maximize a lower bound $\mathcal{L}(\thetab,\phib;\vx)$ on the marginal likelihood $\log p_\thetab(\vx)$,
\begin{equation}
    \log p_\thetab(\vx) = \log \int p_\thetab(\vx,\vz) \, \mathrm{d}\vz  \geq \int q_\phib(\vz|\vx) \log \frac{p_\thetab(\vx|\vz)p(\vz)}{q_\phib(\vz|\vx)}  \, \mathrm{d}\vz \triangleq \mathcal{L}(\thetab,\phib;\vx)  \enspace , \label{eq: elbo}
\end{equation}
where $\thetab$ are parameters of the generative model and $q_\phib(\vz|\vx)$ is the variational approximation to the true posterior.
Graphical illustrations of the models can be seen in \cref{fig: autoregressive vrnn srnn cwvae 1L graphs} and appendix~\cref{app: additional graphical models}.

@@

\paragraph{Results}
We present the model likelihoods for $s=1$ and $64$ in \cref{tab: likelihoods timit} and for $s=256$ in appendix~\cref{app: additional likelihood results appendix}. It is clear that deterministic autoregressive models are superior to LVMs at $s=1$, but inferior to them at $s=64$ and $256$. For strides $s>1$, previous work has attributed the inferior performance of autoregressive models without latent variables to the ability of LVMs to model intra-step correlations \parencite{lai_re-examination_2019}. It should be noted that at $s=1$, the STCN was numerically unstable and that the CW-VAE was computationally infeasible to train.
%
Surprisingly, the simple LSTM performs almost on par with WaveNet. The autoregressive STCN is the best-performing LVM. For both CW-VAE and STCN, increasing the number of latent variables in the hierarchy improves performance, similar to results in the image domain. 
%The best likelihoods correspond to about a 30\% reduction in bit rate which is inferior to FLAC indicating that significant gains in likelihood might be possible. 
The best performing LVMs, STCN and CW-VAE, are not yet scalable to $s=1$ resolution where the highest likelihoods are achieved. Hence, LVMs may be able to outperform autoregressive models at $s=1$ in the future.

\paragraph{Phoneme recognition}
Although the likelihood is a practical metric for model comparison, a high likelihood does not guarantee that a model has learned useful representations \parencite{huszar_is_2017}. To assess the usefulness of the representations, we train a model to recognize phonemes on the TIMIT data. We compare the phoneme error rate (PER) of the model when using input representations obtained from different unsupervised models, trained on the full TIMIT dataset (3.7h) at $s=64$, as in \cref{tab: likelihoods timit}, and when using a Mel spectrogram (80 filterbanks, hop length 64, window size 128). The recognition model is a three-layered bidirectional LSTM with 256 hidden units. It is trained with the connectionist temporal classification (CTC) loss \parencite{graves_connectionist_2006}. We report the PER in \cref{tab: phoneme recognition (PER)}.

>>>>

        \edge[]{d_t}{z_t};
        \edge[bend right]{d_t}{x_t};
        \edge[]{z_tm1}{d_t};
        \edge[]{d_tm1}{d_t};
        \edge[]{z_t}{x_t};
    \end{tikzpicture}
    }
    \end{multicols}
\caption{
Generative models of an LSTM, VRNN, SRNN, STCN and CW-VAE for a single timestep. The STCN and CW-VAE are illustrated with a single latent variable. 
Red arrows indicate purely deterministic paths from the output $\vx_t$ to previous input $\vx_{<t}$ without passing a stochastic node. 
We provide additional graphical illustrations including inference models in \cref{app: additional graphical models}.
}
\label{fig: autoregressive vrnn srnn cwvae 1L graphs}
\end{figure*}

\paragraph{Scope and related work} 
At a high level, this benchmark brings order to LVM model comparisons for speech and also provides useful reference implementations of the models\footnote{\texttt{ \href{https://github.com/JakobHavtorn/benchmarking-lvms}{github.com/JakobHavtorn/benchmarking-lvms}}}. 
An extended version of this paper is available at \href{https://arxiv.org/abs/2202.12707}{arXiv:2202.12707}.
LVMs based on the VAE are of interest due to their ability to learn an approximate posterior distribution over latent variables. This makes them useful beyond generation for e.g. semi"=supervised learning \parencite{kingma_semi-supervised_2014} and anomaly detection \parencite{havtorn_hierarchical_2021}. 
In this paper, we focus on the VRNN \parencite{chung_recurrent_2015}, SRNN \parencite{fraccaro_sequential_2016} and STCN \parencite{aksan_stcn_2019} since they are similar to the original VAE framework. 
We exclude some related work from the benchmark. Specifically, the FH-VAE \parencite{hsu_unsupervised_2017} adds a global latent variable but we exclude it due to it's segmentation of the training data. We also exclude Z-forcing \parencite{goyal_zforcing_2017} which resembles the SRNN, but deviates from maximum likelihood training by using an auxiliary loss. We exclude the VQ-VAE \parencite{oord_neural_2018} which is a hybrid LVM and autoregressive model, since it uses a quantized latent space. Finally, we exclude the Stochastic WaveNet \parencite{lai_stochastic_2018} which is similar, but inferior, to the STCN \parencite{aksan_stcn_2019}.

@@

\section{Models}
\paragraph{Sequential deep latent variable models}
The models we consider are all sequential deep latent variable models trained with variational inference and the reparameterization trick \parencite{kingma_autoencoding_2014}.
The input is a variable-length sequence $\vx = \vx_{1:T} = \left(\vx_1, \vx_2, \dots, \vx_{T}\right)$ with $\vx_t\in\mathbb{R}^{D_x}$. 
The models first encode $\vx_{1:T}$ to a temporal posterior distribution $q(\vz|\vx)$ and sample a latent representation $\vz_{1:T}$ where $\vz_t\in\mathbb{R}^{D_z}$. This is then used to reconstruct the input while the posterior distribution is regularized to be close to a prior distribution $p(\vz_t|\cdot)$ where the dot indicates that it may depend on latent and observed variables at previous timesteps, $\vz_{<t}$ and $\vx_{<t}$ where $\vz_{<t} := \left( \vz_1, \vz_2, \dots, \vz_{t-1} \right)$. 
The models are trained to maximize a lower bound $\mathcal{L}(\thetab,\phib;\vx)$ on the marginal likelihood $\log p_\thetab(\vx)$,
\begin{equation}
    \log p_\thetab(\vx) = \log \int p_\thetab(\vx,\vz) \, \mathrm{d}\vz  \geq \int q_\phib(\vz|\vx) \log \frac{p_\thetab(\vx|\vz)p(\vz)}{q_\phib(\vz|\vx)}  \, \mathrm{d}\vz \equiv \mathcal{L}(\thetab,\phib;\vx)  \enspace , \label{eq: elbo}
\end{equation}
where $\thetab$ are parameters of the generative model and $q_\phib(\vz|\vx)$ is the variational approximation to the true posterior.
Graphical illustrations of the models can be seen in \cref{fig: autoregressive vrnn srnn cwvae 1L graphs} and appendix~\cref{app: additional graphical models}.

@@

\paragraph{Results}
We present the model likelihoods for $s=1$ and $64$ in \cref{tab: likelihoods timit} and for $s=256$ in appendix~\cref{app: additional likelihood results appendix}. It is clear that deterministic autoregressive models are superior to LVMs at $s=1$, but inferior to them at $s=64$ and $256$. For strides $s>1$, previous work has attributed the inferior performance of autoregressive models without latent variables to the ability of LVMs to model intrastep correlations \parencite{lai_re-examination_2019}. It should be noted that at $s=1$, the STCN was numerically unstable and that the CW-VAE was computationally infeasible to train.
%
Surprisingly, the simple LSTM performs almost on par with WaveNet. The autoregressive STCN is the best-performing LVM. For both CW-VAE and STCN, increasing the number of latent variables in the hierarchy improves performance, similar to results in the image domain. 
%The best likelihoods correspond to about a 30\% reduction in bit rate which is inferior to FLAC indicating that significant gains in likelihood might be possible. 
The best performing LVMs, STCN and CW-VAE, are not yet scalable to $s=1$ resolution where the highest likelihoods are achieved. Hence, LVMs may be able to outperform autoregressive models at $s=1$ in the future.

\paragraph{Phoneme recognition}
Although the likelihood is a practical metric for model comparison, a high likelihood does not guarantee that a model has learned useful representations \parencite{huszar_is_2017}. To assess the usefulness of the representations, we train a model to recognize phonemes on the TIMIT data. We compare the phoneme error rate (PER) of the model when using input representations obtained from different unsupervised models, trained on the full TIMIT dataset (3.7h) at $s=64$, as in \cref{tab: likelihoods timit}, and when using a Mel spectrogram (80 filter banks, hop length 64, window size 128). The recognition model is a three-layered bidirectional LSTM with 256 hidden units. It is trained with the connectionist temporal classification (CTC) loss \parencite{graves_connectionist_2006}. We report the PER in \cref{tab: phoneme recognition (PER)}.

### 2023-10-14 chapters/paper_hierarchical.tex
Fix many spelling and language errors

To shift the ELBO from primarily being based on the approximate posterior of the lowest latent variables to instead focus on the conditional prior, \textcite{maaloe_biva_2019} introduced slightly different likelihood lower bound defined as
\begin{equation}\label{eq_hierarchical:biva->k}
    \mathcal{L}^{>k} = \mathbb{E}_{p_\thetab(\zb_{\leq k}|\zb_{>k}) q_\phib(\zb_{>k}|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p_\thetab(\zb_{>k})}{q_\phib(\zb_{>k}|\xb)} \right]
\end{equation}
where $k\in\{0,1,\dots,L\}$ (see \cref{sec_hierarchical:derivation_L_geq_k} for the derivation).
We note that $\mathcal{L}^{>0}$ is the regular ELBO (\cref{eq_hierarchical:elbo}) and that empirically we always observe that $\mathcal{L}\geq\mathcal{L}^{>k} \, \forall \, k$ although this need not hold in general.
The core idea behind this variation on the ELBO is to sample the $k$ lowest latent variables from the conditional prior $\zb_1,\dots,\zb_l \sim p_\thetab(\zb_{\leq k}|\zb_{>k})$ and only the $L-k$ highest from the approximate posterior $\zb_{k+1},\dots,\zb_L \sim q_\phib(\zb_{>k}|\xb)$.
Importantly, this has the effect that the data likelihood $p(\xb|\zb)$ is dependent on the approximate posterior through a latent variable $\zb_{k+1}$ different from $\zb_1$ for all $k \geq 1$.
Thereby, the likelihood can be evaluated with a reconstruction from each of the latent variables $\zb_k$ of the hierarchical VAE.
Hence, we can now test how well the input $\xb$ is reconstructed from each latent variable.
The notation $\mathcal{L}^{>k}$ highlights that for latent variables $\zb_{>k}$, the bound is the regular ELBO while for the latent variables $\zb_{\leq k}$, the bound is evaluated using the (conditional) prior rather than the approximate posterior as the proposal distribution.

@@

\subsection{Likelihood-based OOD detection}
% \begin{sidewaysfigure}
\begin{figure}
    %\captionstyle{centerlast}
    \centering
    \begin{subfigure}[l]{0.48\columnwidth}
        \centering
        \includegraphics[width=1\columnwidth]{paper_hierarchical/densities-FashionMNIST-test-MNIST-test-bpp-k-0_sohau.pdf}
        \caption{}
        \label{fig_hierarchical:FMNIST-elbo-k0}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[c]{0.48\columnwidth}
        \centering
        \includegraphics[width=1\columnwidth]{paper_hierarchical/densities-FashionMNIST-test-MNIST-test-bpp-k-2_sohau.pdf}
        \caption{}
        \label{fig_hierarchical:FMNIST-elbo-k2}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[r]{0.48\columnwidth}
        \centering
        \includegraphics[width=1\columnwidth]{paper_hierarchical/densities-FashionMNIST-test-MNIST-test-LLR-0-1_sohau.pdf}
        \caption{}
        \label{fig_hierarchical:FMNIST-llr}
    \end{subfigure}
    \caption[Empirical densities of FashionMNIST (in-distribution) and MNIST (OOD) using the raw likelihood and $\mathcal{L}^{>2}$ bound.]{%
    Empirical densities of FashionMNIST (in-distribution) and MNIST (OOD) using the raw likelihood \subref{fig_hierarchical:FMNIST-elbo-k0}, the $\mathcal{L}^{>2}$ bound \subref{fig_hierarchical:FMNIST-elbo-k2} and the $LLR^{>1}$ score \subref{fig_hierarchical:FMNIST-llr}. All densities are computed using the HVAE model.
    For the regular likelihood MNIST is very clearly more likely on average than the FashionMNIST test data while with the $\mathcal{L}^{>2}$ bound separation is better but significant overlap remains.
    The $LLR^{>1}$ provides a high degree of separation. Likelihoods are reported in units of the natural log of the number of bits per dimension.
    }
    \label{fig_hierarchical:FMNIST-ood-densities}
\end{figure}
% \end{sidewaysfigure}

>>>>

To shift the ELBO from primarily being based on the approximate posterior of the lowest latent variables to instead focus on the conditional prior, \textcite{maaloe_biva_2019} introduced slightly different likelihood lower bound defined as
\begin{equation}\label{eq_hierarchical:biva->k}
    \mathcal{L}^{>k} = \mathbb{E}_{p_\thetab(\zb_{\leq k}|\zb_{>k}) q_\phib(\zb_{>k}|\xb)} \left[ \log \frac{p_\thetab(\xb|\zb)p_\thetab(\zb_{>k})}{q_\phib(\zb_{>k}|\xb)} \right]
\end{equation}
where $k\in\{0,1,\dots,L\}$ (see \cref{sec_hierarchical:derivation_L_geq_k} for the derivation).
We note that $\mathcal{L}^{>0}$ is the regular ELBO (\cref{eq_hierarchical:elbo}) and that empirically we always observe that $\mathcal{L}\geq\mathcal{L}^{>k} \, \forall \, k$, although this need not hold in general.
The core idea behind this variation on the ELBO is to sample the $k$ lowest latent variables from the conditional prior $\zb_1,\dots,\zb_l \sim p_\thetab(\zb_{\leq k}|\zb_{>k})$ and only the $L-k$ highest from the approximate posterior $\zb_{k+1},\dots,\zb_L \sim q_\phib(\zb_{>k}|\xb)$.
Importantly, this has the effect that the data likelihood $p(\xb|\zb)$ is dependent on the approximate posterior through a latent variable $\zb_{k+1}$ different from $\zb_1$ for all $k \geq 1$.
Thereby, the likelihood can be evaluated with a reconstruction from each of the latent variables $\zb_k$ of the hierarchical VAE.
Hence, we can now test how well the input $\xb$ is reconstructed from each latent variable.
The notation $\mathcal{L}^{>k}$ highlights that for latent variables $\zb_{>k}$, the bound is the regular ELBO while for the latent variables $\zb_{\leq k}$, the bound is evaluated using the (conditional) prior rather than the approximate posterior as the proposal distribution.

@@

\subsection{Likelihood-based OOD detection}
% \begin{sidewaysfigure}
\begin{figure}
    %\captionstyle{centerlast}
    \centering
    \begin{subfigure}[l]{0.48\columnwidth}
        \centering
        \includegraphics[width=1\columnwidth]{paper_hierarchical/densities-FashionMNIST-test-MNIST-test-bpp-k-0_sohau.pdf}
        \caption{}
        \label{fig_hierarchical:FMNIST-elbo-k0}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[c]{0.48\columnwidth}
        \centering
        \includegraphics[width=1\columnwidth]{paper_hierarchical/densities-FashionMNIST-test-MNIST-test-bpp-k-2_sohau.pdf}
        \caption{}
        \label{fig_hierarchical:FMNIST-elbo-k2}
    \end{subfigure}
    % \hfill
    \begin{subfigure}[r]{0.48\columnwidth}
        \centering
        \includegraphics[width=1\columnwidth]{paper_hierarchical/densities-FashionMNIST-test-MNIST-test-LLR-0-1_sohau.pdf}
        \caption{}
        \label{fig_hierarchical:FMNIST-llr}
    \end{subfigure}
    \caption[Empirical densities of FashionMNIST (in-distribution) and MNIST (OOD) using the raw likelihood and $\mathcal{L}^{>2}$ bound.]{ Empirical densities of FashionMNIST (in-distribution) and MNIST (OOD) using the raw likelihood \subref{fig_hierarchical:FMNIST-elbo-k0}, the $\mathcal{L}^{>2}$ bound \subref{fig_hierarchical:FMNIST-elbo-k2} and the $LLR^{>1}$ score \subref{fig_hierarchical:FMNIST-llr}. All densities are computed using the HVAE model. For the regular likelihood MNIST is very clearly more likely on average than the FashionMNIST test data while with the $\mathcal{L}^{>2}$ bound separation is better but significant overlap remains. The $LLR^{>1}$ provides a high degree of separation. Likelihoods are reported in units of the natural log of the number of bits per dimension.
    }
    \label{fig_hierarchical:FMNIST-ood-densities}
\end{figure}
% \end{sidewaysfigure}

### 2023-10-14 chapters/paper_modelagnostic.tex
Fix many spelling and language errors

Let us discuss our choice of combining the score statistic and the typicality test. We will try to look in which situations one of the test fails and the other works and vice versa. Both examples assume that the in-distribution data follows a $\mathcal{N}(0,I_D)$ distribution, and that the correct model has been learned by fitting $(\mathcal{N}(\theta,I_D))_{\theta \in \mathbb{R}^D}$ via maximum-likelihood. Even in this simple setting with no model misspecification, we will see that the two statistics that we consider may have very different strengths.

@@

%%%%%%%%%%%%%%%%%%%%%%% FIGURE
\begin{figure}[tb]
    \centering
    \includegraphics[scale=0.6]{graphics/paper_modelagnostic/fashion_hvae_type1_correct_font.pdf}
    \caption[Type I and Type II errors versus the significance level $\alpha$ on the combination values]{Type I (probability of an inlier to be classified as outlier) and Type II (probability of an outlier to be considered as inlier) errors versus the significance level $\alpha$ on the combination values. By using Benjamini-Hochberg correction, we get that the Type I error stays below identity line.}
    \label{fig_modelagnostic:type1}
\end{figure}

>>>>

Let us discuss our choice of combining the score statistic and the typicality test. We will try to look in which situations one of the test fails and the other works and vice versa. Both examples assume that the in-distribution data follows an $\mathcal{N}(0,I_D)$ distribution, and that the correct model has been learned by fitting $(\mathcal{N}(\theta,I_D))_{\theta \in \mathbb{R}^D}$ via maximum-likelihood. Even in this simple setting with no model misspecification, we will see that the two statistics that we consider may have very different strengths.

@@

%%%%%%%%%%%%%%%%%%%%%%% FIGURE
\begin{figure}[tb]
    \centering
    \includegraphics[scale=0.6]{graphics/paper_modelagnostic/fashion_hvae_type1_correct_font.pdf}
    \caption[Type I and Type II errors versus the significance level $\alpha$ on the combination values.]{ Type I (probability of an inlier to be classified as outlier) and Type II (probability of an outlier to be considered as inlier) errors versus the significance level $\alpha$ on the combination values. By using Benjamini-Hochberg correction, we get that the Type I error stays below identity line.}
    \label{fig_modelagnostic:type1}
\end{figure}
